---
title: "Univariate Analysis: Bike-Sharing Dataset"
author: "Statistical Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
```

# Setup and Data Loading

```{r libraries}
library(corrplot)
library(GGally)
library(moments)
```

```{r load-data}
data <- read.csv("hour.csv")

# Re-labeling categorical variables for better interpretability
data$season_lab <- factor(data$season, labels = c("Spring", "Summer", "Fall", "Winter"))
data$weather_lab <- factor(data$weathersit, labels = c("Clear", "Mist", "Light Snow/Rain", "Heavy Rain"))

# Dataset dimensions
cat("Observations:", nrow(data), "\nVariables:", ncol(data))
```
```{r data-gap-analysis}
library(dplyr)
library(lubridate)

# Ensure the date column is in Date format
data$dteday <- as.Date(data$dteday)

# Identify the start and end of the observations
start_date <- min(data$dteday)
end_date <- max(data$dteday)

# Create a theoretical complete timeline (every hour for every day)
theoretical_timeline <- expand.grid(
  dteday = seq.Date(start_date, end_date, by = "day"),
  hr = 0:23
)

# Identify missing records (gaps) by joining the actual data with the timeline
check_gap <- theoretical_timeline %>%
  left_join(data, by = c("dteday", "hr")) %>%
  filter(is.na(cnt)) # Filter only records that do not exist in your dataset

# Summarize the number of missing hours per day
gap_per_day <- check_gap %>%
  group_by(dteday) %>%
  summarise(missing_hours = n()) %>%
  arrange(desc(missing_hours))

# Display the results
gap_per_day
```

**Note**: The dataset contains hourly bike rental data from 2011-01-01 to 2012-12-31. A completeness check reveals some missing hours on certain days. After investigations, these gaps are caused by natural disaster events (e.g., hurricane, snow storm) leading to station closures.

---

# Temporal Categorical Variables

## Season

```{r season}
# Frequency distribution
freq_season <- data.frame(table(data$season_lab)) %>%
  rename(season = Var1, frequency = Freq) %>%
  mutate(proportion = frequency / sum(frequency))

print(freq_season)

# Visualization with ggplot2
p_season <- ggplot(freq_season, aes(x = season, y = frequency, fill = season)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.5) +
  geom_text(aes(label = round(proportion, 3)), vjust = -0.5, size = 3.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Season Distribution",
       x = "Season",
       y = "Frequency",
       fill = "Season") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

print(p_season)

# Chi-square test for uniformity
chisq.test(freq_season$frequency)
```

**Interpretation**: Dataset is well-balanced across the four seasons (~25% each).

## Year

```{r year}
freq_year <- data.frame(table(data$yr)) %>%
  rename(year = Var1, frequency = Freq) %>%
  mutate(year = factor(year, labels = c("2011", "2012")),
         proportion = frequency / sum(frequency))

print(freq_year)

# Visualization 
p_year <- ggplot(freq_year, aes(x = year, y = frequency, fill = year)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.5) +
  geom_text(aes(label = round(proportion, 3)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("2011" = "skyblue", "2012" = "salmon")) +
  labs(title = "Yearly Distribution",
       x = "Year",
       y = "Frequency",
       fill = "Year") +
  theme_minimal() +
  theme(legend.position = "none")

print(p_year)
```

## Month

```{r month}
freq_month <- table(data$mnth)
print(freq_month)

barplot(freq_month, main = "Monthly Distribution", 
        col = "lightgreen", xlab = "Month", ylab = "Frequency")

# Uniformity test
chisq.test(freq_month)
```

## Hour

```{r hour}
freq_hour <- data.frame(table(data$hr)) %>%
  rename(hour = Var1, frequency = Freq) %>%
  mutate(hour = as.numeric(hour) - 1,
         proportion = frequency / sum(frequency))

print(summary(freq_hour$hour))

# Visualization
p_hour <- ggplot(freq_hour, aes(x = hour, y = frequency, fill = frequency)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3) +
  scale_fill_gradient(low = "lavender", high = "purple") +
  scale_x_continuous(breaks = seq(0, 23, by = 2)) +
  labs(title = "Hourly Distribution",
       x = "Hour (0-23)",
       y = "Frequency",
       fill = "Frequency") +
  theme_minimal() +
  theme(legend.position = "right")

print(p_hour)

# Uniformity test
chisq.test(freq_hour$frequency)
```


## Weekday

```{r weekday}
freq_weekday <- data.frame(table(data$weekday)) %>%
  rename(weekday = Var1, frequency = Freq) %>%
  mutate(weekday = factor(as.numeric(weekday) - 1, 
                          labels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")),
         proportion = frequency / sum(frequency))

print(freq_weekday)

# Visualization
p_weekday <- ggplot(freq_weekday, aes(x = weekday, y = frequency, fill = weekday)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.5) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Weekday Distribution",
       x = "Weekday",
       y = "Frequency",
       fill = "Weekday") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

print(p_weekday)

chisq.test(freq_weekday$frequency)
```

---

## Holiday

```{r holiday}
freq_holiday <- data.frame(table(data$holiday)) %>%
  rename(holiday = Var1, frequency = Freq) %>%
  mutate(holiday = factor(holiday, labels = c("Non-Holiday", "Holiday")),
         proportion = frequency / sum(frequency))

print(freq_holiday)

# Visualization
p_holiday <- ggplot(freq_holiday, aes(x = holiday, y = frequency, fill = holiday)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.5) +
  geom_text(aes(label = paste0(round(proportion * 100, 1), "%")), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Non-Holiday" = "lightblue", "Holiday" = "tomato")) +
  labs(title = "Holiday Distribution",
       x = "Holiday Status",
       y = "Frequency",
       fill = "Status") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p_holiday)
```

**Note**: Proportion reflects the rarity of holidays in the calendar.

## Working Day

```{r workingday}
freq_working <- data.frame(table(data$workingday)) %>%
  rename(workingday = Var1, frequency = Freq) %>%
  mutate(workingday = factor(workingday, labels = c("Non-Working", "Working")),
         proportion = frequency / sum(frequency))

print(freq_working)

# Visualization
p_working <- ggplot(freq_working, aes(x = workingday, y = frequency, fill = workingday)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.5) +
  geom_text(aes(label = paste0(round(proportion * 100, 1), "%")), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Non-Working" = "lightcyan", "Working" = "steelblue")) +
  labs(title = "Working Day Distribution",
       x = "Working Day Status",
       y = "Frequency",
       fill = "Status") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p_working)
```

**Interpretation**: About 70% of observations represent working days, consistent with standard calendar distribution.

## Weather Situation

```{r weather-distribution}
library(ggplot2)

# Visualization of Weather Conditions
ggplot(data, aes(x = weather_lab, fill = weather_lab)) +
  geom_bar(color = "black", show.legend = FALSE) +
  stat_count(geom = "text", 
             aes(label = after_stat(count)), 
             vjust = -0.5,
             size = 4.5, 
             fontface = "bold") +
  scale_fill_manual(values = c("Clear" = "yellow", 
                               "Mist" = "lightgrey", 
                               "Light Snow/Rain" = "darkgrey", 
                               "Heavy Rain" = "black")) +
  
  labs(title = "Weather Condition Distribution",
       x = "Weather Condition",
       y = "Frequency") +

  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +

  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.major.x = element_blank())
```

**Interpretation**: Most rentals occur during "Clear/Few Clouds" (Category 1). Category 4 (Heavy Rain/Snow) is extremely rare (n = 3), representing a rare event. Sharp class imbalance present.

---

# Continuous Variables (Normalized)

## Temperature (temp)

```{r temp}
# Descriptive statistics
cat("Temperature Statistics:\n")
summary(data$temp)
cat("\nStd Dev:", sd(data$temp), "\n")
cat("Coefficient of Variation:", sd(data$temp)/mean(data$temp), "\n")

# Visualization
par(mfrow = c(1, 3))
hist(data$temp, main = "Temperature Histogram", 
     col = "seagreen", border = "white", xlab = "Temp (normalized)", 
     probability = TRUE, breaks = 30)
lines(density(data$temp), col = "red", lwd = 2)

boxplot(data$temp, main = "Temperature Boxplot", col = "gold")

qqnorm(data$temp, main = "Q-Q Plot Temperature")
qqline(data$temp, col = "red")
par(mfrow = c(1, 1))

# Normality test
shapiro.test(sample(data$temp, 5000))
```

**Interpretation**: Roughly bimodal distribution, reflecting seasonal nature of data.

## Feeling Temperature (atemp)

```{r atemp}
cat("Feeling Temperature Statistics:\n")
summary(data$atemp)
cat("\nStd Dev:", sd(data$atemp), "\n")
cat("Coefficient of Variation:", sd(data$atemp)/mean(data$atemp), "\n")

par(mfrow = c(1, 3))
hist(data$atemp, main = "Feel Temperature Histogram", 
     col = "coral", border = "white", xlab = "ATemp (normalized)", 
     probability = TRUE, breaks = 30)
lines(density(data$atemp), col = "blue", lwd = 2)

boxplot(data$atemp, main = "Feel Temperature Boxplot", col = "lightblue")

qqnorm(data$atemp, main = "Q-Q Plot Feel Temperature")
qqline(data$atemp, col = "blue")
par(mfrow = c(1, 1))

shapiro.test(sample(data$atemp, 5000))
```

**Critical Note**: Extremely correlated with `temp` (r > 0.98), indicating redundant information.

## Humidity (hum)

```{r humidity}
cat("Humidity Statistics:\n")
summary(data$hum)
cat("\nStd Dev:", sd(data$hum), "\n")
cat("Coefficient of Variation:", sd(data$hum)/mean(data$hum), "\n")

par(mfrow = c(1, 3))
hist(data$hum, main = "Humidity Histogram", 
     col = "skyblue", border = "white", xlab = "Humidity (normalized)", 
     probability = TRUE, breaks = 30)

boxplot(data$hum, main = "Humidity Boxplot", col = "lightcyan")

qqnorm(data$hum, main = "Q-Q Plot Humidity")
qqline(data$hum, col = "darkblue")
par(mfrow = c(1, 1))

shapiro.test(sample(data$hum, 5000))

# Skewness
cat("\nSkewness:", moments::skewness(data$hum))
```
**Interpretation**: Slightly right-skewed distribution. Most records show humidity levels between 40% and 80%. Weak negative correlation with bike rentals (r ≈ -0.10), as very high humidity often correlates with rain or discomfort.

## Wind Speed (windspeed)

```{r windspeed}
cat("Wind Speed Statistics:\n")
summary(data$windspeed)
cat("\nStd Dev:", sd(data$windspeed), "\n")
cat("Coefficient of Variation:", sd(data$windspeed)/mean(data$windspeed), "\n")

par(mfrow = c(1, 3))
hist(data$windspeed, main = "Wind Speed Histogram", 
     col = "lightgreen", border = "white", xlab = "Wind Speed (normalized)", 
     probability = TRUE, breaks = 30)

boxplot(data$windspeed, main = "Wind Speed Boxplot", col = "palegreen")

qqnorm(data$windspeed, main = "Q-Q Plot Wind Speed")
qqline(data$windspeed, col = "darkgreen")
par(mfrow = c(1, 1))

shapiro.test(sample(data$windspeed, 5000))
```

**Interpretation**: Left-skewed distribution. Significant number of records show "0" values, potentially representing near-calm conditions or sensor thresholds. High wind speeds are rare and typically act as a deterrent for casual users.

---

# Count Variables (Response Variables)

## Casual Users

```{r casual}
cat("Casual Users Statistics:\n")
summary(data$casual)
cat("\nStd Dev:", sd(data$casual), "\n")
cat("Coefficient of Variation:", sd(data$casual)/mean(data$casual), "\n")

par(mfrow = c(1, 2))
hist(data$casual, main = "Casual Users Distribution", 
     col = "orange", breaks = 50, xlab = "Count")

boxplot(data$casual, main = "Casual Users Boxplot", col = "peachpuff")
par(mfrow = c(1, 1))

# Skewness and kurtosis
cat("\nSkewness:", moments::skewness(data$casual), "\n")
cat("Kurtosis:", moments::kurtosis(data$casual), "\n")
```

**Interpretation**: Right-skewed count distribution. Casual users show sporadic, lower-volume behavior.

## Registered Users

```{r registered}
cat("Registered Users Statistics:\n")
summary(data$registered)
cat("\nStd Dev:", sd(data$registered), "\n")
cat("Coefficient of Variation:", sd(data$registered)/mean(data$registered), "\n")

par(mfrow = c(1, 2))
hist(data$registered, main = "Registered Users Distribution", 
     col = "purple", breaks = 50, xlab = "Count")

boxplot(data$registered, main = "Registered Users Boxplot", col = "plum")
par(mfrow = c(1, 1))

cat("\nSkewness:", moments::skewness(data$registered), "\n")
cat("Kurtosis:", moments::kurtosis(data$registered), "\n")
```

**Interpretation**: Registered users show much higher volume and drive overall dataset patterns.

## Total Count (cnt - Target Variable)

```{r total-count-analysis}
# Display summary statistics for the total count
cat("Total Count Statistics:\n")
summary(data$cnt)
cat("\nStd Dev:", sd(data$cnt), "\n")
cat("Coefficient of Variation:", sd(data$cnt)/mean(data$cnt), "\n")

# Set up a side-by-side plot layout
par(mfrow = c(1, 2))
hist(data$cnt, main = "Total Count Distribution", 
     col = "firebrick", breaks = 50, xlab = "Count")

boxplot(data$cnt, main = "Total Count Boxplot", col = "salmon")

# Reset plot layout to default
par(mfrow = c(1, 1))

# Display shape statistics
cat("\nSkewness:", moments::skewness(data$cnt), "\n")
cat("Kurtosis:", moments::kurtosis(data$cnt), "\n")

# Verify the additive relationship: total = casual + registered
cat("\nVerification cnt = casual + registered:\n")
all.equal(data$cnt, data$casual + data$registered)
```

```{r}
library(ggplot2)

# Calculate reference statistics for plot annotations
mu <- mean(data$cnt)
med <- median(data$cnt)

ggplot(data, aes(x = cnt)) +
  # 1. Histogram calculated as density
  # 'after_stat(density)' is key to aligning it with the density curve
  geom_histogram(aes(y = after_stat(density)), bins = 50, 
                 fill = "#4e79a7", color = "white", alpha = 0.6) +
  
  # 2. Smooth density curve
  geom_density(color = "#2c3e50", linewidth = 1.2) +
  
  # 3. Vertical lines for Mean and Median
  geom_vline(aes(xintercept = mu, color = "Mean"), linetype = "dashed", linewidth = 1) +
  geom_vline(aes(xintercept = med, color = "Median"), linetype = "solid", linewidth = 1) +
  
  # 4. Color palette and legend configuration
  scale_color_manual(name = "Statistics:", 
                     values = c("Mean" = "#e67e22", "Median" = "#27ae60")) +
  
  # 5. Axis labels
  labs(
       x = "Number of Rentals (cnt)",
       y = "Density") +
  
  # 6. Professional styling and theme adjustments
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, color = "#2c3e50"),
    plot.subtitle = element_text(size = 13, color = "#7f8c8d"),
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

- **Distribution**: Strong right-skewed distribution. Most observations concentrated in lower range (0-200 rentals per hour), with long tail extending towards higher values.
- **Overdispersion**: Variance >> mean, indicating overdispersion typical of count data.
Logarithmic transformation might be beneficial for linear modeling to normalize residuals. Alternatively, Poisson or Negative Binomial regression more appropriate than standard OLS.

---

